---
title: "EEMB 148 - HW 2"
subtitle: "Algal Assessment"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: sandstone
runtime: shiny_prerendered
description: "Algal analysis for HW 2."
---

```{r setup, include=FALSE}

# Necessary packages:

library(learnr)
library(tidyverse)
library(agricolae)

# Necessary data:

afdm <- read_csv('afdm.csv') %>% 
  select(site, transect, afdm)

chla <- read_csv('chla.csv') %>% 
  select(site, transect, chla) 

# Data wrangling:

algae <- full_join(afdm, chla) %>% 
  select(-transect)

# Tutorial options:

tutorial_options(exercise.timelimit = 10)

```

<style type="text/css">
.table {

    width: 500px;

}
</style>

## Study System

### Santa Ana River

The Santa Ana River flows through southern California, and its drainage basin includes extensive agricultural development.

This study surveyed a stretch of the river near Riverside, which was affected by several agricultural alterations:

- **Riparian vegetation** was removed to increase crop production area.
- Crops were planted **directly adjacent** to the river.
- **Pesticide and fertilizer** treatments were applied on a bi-monthly basis.

<br>

### Study Design

A research team surveyed this stretch of the Santa Ana River in order to quantify changes in algal biomass associated with agricultural development. Within a 1000-m stretch of river spanning **both upstream and downstream of the agricultural area,** they measured Ash Free Dry Mass (AFDM) and Chlorophyll-*a* (Chl-*a*) concentrations.

**Six sites** were chosen at random. Each site was named after a type of mammal that is common in riparian habitat: beaver, mouse, shrew, vole, weasel, and woodrat. At each site, **five replicates (measurements)** were taken for each parameter. Therefore, each parameter has $6*5=30$ total measurements.

**Experimental Design:**

- *Predictor Variable:* location (site)
- *Response Variables:* AFDM and chl-*a*

<br>

### **Task:** Statistical Sleuthing

Your goal is to determine which of the six sites were located downstream of agricultural development, as indicated by changes in algal biomass.

You should be able to back up all claims by reporting statistical evidence.

***Graded Assignment:*** *Homework #2 is formatted as a "quiz" on Gauchospace. Some of the questions ask you to use articles from the scientific literature, which you will be able to find on Gauchospace.*

## AFDM and Chl-a Data

The research team measured two algal parameters:

| Variable Name | Algal Parameter | Units |
| :- | :- | :- |
| `afdm` | Ash Free Dry Mass (AFDM) | mg/L |
| `chla` | Chlorophyll-*a* (Chl-*a*) | mg/L |

Here is the algal data presented in a table. There are 30 rows, since there are 30 measurements for each parameter.

```{r chemical-data, echo=FALSE}
algae
```

## -- ANOVA

$\bigstar$ **Task:** For both AFDM and Chl-*a*, run an ANOVA to determine whether the mean varies by site.

The variable names are: `afdm` and `chla`.

```{r physical-ANOVA, exercise=TRUE}

# Run an ANOVA to determine whether the mean varies by site

algae_anova <- aov(VARIABLE ~ site,
                  data = algae)
summary(algae_anova)

```

<div id="physical-ANOVA-hint">
**Hint:** Replace the word `VARIABLE` with either `afdm` or `chla`. Press "Run Code". Find the ANOVA p-value labeled `Pr(>F)`.
</div>

## -- Tukey test

$\bigstar$ **Task:** For both AFDM and Chl-*a*, run a Tukey test to determine which sites were different.

The variable names are: `afdm` and `chla`.

```{r physical-Tukey, exercise=TRUE}

# Run a Tukey test using the output from the ANOVA

algae_anova <- aov(VARIABLE ~ site,
                  data = algae)
algae_tukey <- HSD.test(algae_anova, "site")
print(algae_tukey$groups)

```

<div id="physical-Tukey-hint">
**Hint:** Replace the word `VARIABLE` with either `afdm` or `chla`. Find the mean values for each site listed under the name of the variable. Determine which sites are similar based on the letters they share.
</div>

## -- Bar graph

$\bigstar$ **Task:** Visualize the data to determine which sites were affected by the agricultural development.

The variable names are: `afdm` and `chla`.

```{r physical-plot, exercise=TRUE, fig.width=5, fig.height=3}

# Create a summary table for plotting:
algae_summary <- algae %>% 
  group_by(site) %>% 
  summarise(
    mean = mean(VARIABLE),
    sd = sd(VARIABLE)
  )

# Create a graph using the summary table:
ggplot(algae_summary, aes(x = site, y = mean)) +
  geom_col() + 
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.2, position = position_dodge(.9)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  labs(title = "Mean Algal Biomass",
       x = "Site", y = "Algal Biomass")
       
```

<div id="physical-plot-hint">
**Hint:** You need to replace `VARIABLE` in the code **twice**.
</div>

## Sources

**Congratulations, you've reached the end of this tutorial!** Wash your hands and get yourself a snack. (:

Algal data modified from dataset courtesy of Thomas Even.

Tutorial by Kate Culhane (kathrynculhane@ucsb.edu).