---
title: "Introduction to R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Introduction to R and some of its uses.
---

```{r setup, include=FALSE}

# Necessary packages:

library(learnr)
library(tidyverse)
library(lubridate)
library(plotly)

# Data for Goleta rainfall/discharge:

water <- read_csv('goleta_rainfall_discharge.csv') %>% 
  mutate(type = factor(type, levels = c('Rainfall', 'Discharge'))) %>% 
  mutate(date = dmy(date))

water2 <- water %>% 
  pivot_wider(names_from = type, values_from = value) %>% 
  select(date, Rainfall, Discharge) %>% 
  rename(Date = date)

labs <- c("Daily Rainfall (in)", "Average Discharge (cubic ft/s)")
names(labs) <- c("Rainfall", "Discharge")

# Data for Sierra streams:

algae <- read_csv('week2_algae.csv') %>% 
  gather(key = stream_type, value = algal_biomass) %>% 
  mutate(stream_type = case_when(
    stream_type == 'fish' ~ 'Trout',
    stream_type == 'declining' ~ 'Declining',
    TRUE ~ 'No Trout'
  )) %>% 
  mutate(stream_type = factor(stream_type, levels = c('No Trout', 'Declining', 'Trout'))) 

algae2 <- algae %>% 
  filter(stream_type != "Declining") %>% 
  mutate(stream_type = factor(stream_type, levels = c('No Trout', 'Trout')))

algae_wider <- algae %>% 
  mutate(id = rep(1:50,3)) %>% 
  pivot_wider(names_from = stream_type, values_from = algal_biomass) %>% 
  select(-1) %>% 
  select(Trout, Declining, `No Trout`)

algae_anova <- aov(algal_biomass ~ stream_type, data = algae)

# Tutorial options:

tutorial_options(exercise.timelimit = 10)

```

## What is this tutorial?

### Introduction to R for EEMB 148

This tutorial will help you become familiar with R and how we'll be using it for EEMB 148. It is optional: there is *no graded assignment associated with this tutorial.* 

We will be using a similar R tutorial to run statistical tests and create graphs for the homework assignments and case study, but our goal is allow you to tailor this experience to your skill level:

* **If you are an R novice** - Interact with R entirely through the tutorials; you will only need to press a button to create the needed statistical output and graphs. This will be sufficient to earn full points.
* **If you are building up confidence in R** - Start with the code that we provide, and either modify it within the tutorial interface or copy into your own project in R Studio.
* **If you are an R expert** - Feel free to write your own code in R Studio! Necessary datasets will be provided on Gauchospace.

This tutorial should take *less than an hour* to complete.

<br>

### Outline

First we'll explore some fun uses of R:

* Making calculations
* Exploring data
* Running statistical tests
* Creating models
* Creating interactive graphs

Then we'll review how we are using R within EEMB 148:

* Running t-tests
* Running ANOVAs
* Running Tukey tests
* Creating bar graphs

*Note: At any time, you can re-set this tutorial by clicking "Start Over" at the end of the Table of Contents.*

## What is R?

R is a **statistical programming language.** It is particularly useful for anyone who works with data: scientists, healthcare workers, manufacturers, financial advisors... In all of these fields, R is used to clean, explore, analyze, visualize, and present data. In stream ecology, we often use R to run statistical tests, create models, and make figures.

In this course, we will be using R for some simple tasks, including running basic statistical tests and creating bar graphs.

## Uses of R

Let's explore some of the things R can do...

### At its most basic, R is a **calculator:**

*Press the blue "Run Code" button!*

*To re-set the code, press the "Start Over" button.*

*Notes are indicated in green.*

```{r addition, exercise=TRUE}

# Add two numbers together:
48 + 390

```

There are many **operators** besides `+`. A few common ones are:

- `-` subtract
- `*` multiply
- `/` divide

$\bigstar$ **Try this:** Add another operator and number to the code above. Press "Run Code" again to see what happens!

<br>

### We can use R to **explore data:**

The following data represent the daily rainfall (in inches) in Goleta for 2019 as well as the average stream discharge (amount of flowing water in cubic feet per second) in Atascadero Creek. Only the first five lines of data are shown, from the first five days of 2019.

```{r data, echo=FALSE}

head(water2, 5)

```

<br>

We can use R to create a scatterplot of this data, in order to explore the relationship between rainfall and discharge:

```{r scatter_plot, exercise=TRUE, fig.height=3, fig.width=4.5}

# Create a scatterplot:
ggplot(data = water2, # data comes from a dataset named "water2"
       aes (x = Rainfall, # x-axis shows rainfall
            y = Discharge)) + # y-axis shows discharge
  geom_point(color = "darkblue") + # changes color of points 
  theme_classic() + # changes the theme
  labs(title = "Rainfall and Stream Discharge in Goleta, 2019", # graph title
       x = "Daily Rainfall (in)", # x-axis title
       y = "Average Discharge (cubic ft/s)") # y-axis title

```

$\bigstar$ **Try this:** You can change the color of the points by replacing `"darkblue"` with `"turquoise"`. Try a few different colors to see which work!

$\bigstar$ **Try more:** Replace `(color = "darkblue")` with `(color = "darkblue", shape = 2)`. What happens? You can play around with other numbers as well to change your graph!

<br>

### R is a **statistical** powerhouse:

There are many statistical tests we can conduct in R, including tests of correlation. We can quantify the correlation between rainfall and discharge using a **correlation coefficient**. The correlation coefficient shows how strongly two variables are correlated, and ranges from -1 to 1:

- -1 indicates a strong negative linear relationship
- 0 indicates no relationship
- 1 indications a strong positive linear relationship

```{r correlation_test, exercise=TRUE}

# Calculate the correlation between rainfall and discharge:
cor(water2$Rainfall, water2$Discharge)

```

$\bigstar$ **Think:** Does this correlation coefficient indicate a strong positive or negative relationship?

<br>

### You can also create **models:**

One type of model is a **linear model**, which shows how two variables are related by using a straight line. The model results tell us the coefficients and intercept of that line, as well as a **p-value** associated with significance.

```{r linear_model, exercise=TRUE}

# Create a linear model:
water_linear_model <- lm(data = water2, Discharge ~ Rainfall)

# Summarize the linear model:
summary(water_linear_model)

```

$\bigstar$ **Find:** Look for the column labeled `Pr(>|t|)`.This is the p-value! Find the p-value associated with "Rainfall"...it is very very small and is followed by three asterisks.

**Based on this very short analysis, we can see that rainfall has a *significant* effect on discharge (p < 0.05)!**

<br>

Now we can add our linear model to the scatterplot we made earlier:

```{r lm_graph, exercise=TRUE, fig.height=3, fig.width=4.5}

# Graph of the linear model:
ggplot(data = water2, # data is from a dataset named "water2"
       aes (x = Rainfall, # x-axis is Rainfall
            y = Discharge)) + # y-axis is Discharge
  geom_smooth(method = lm, # adds a line for the linear model
              color = "red") + # changes the color of the line
  geom_point(color = "darkblue") + # changes the color of the points
  theme_classic() + # changes the theme
  labs(title = "Linear Model of Rainfall and Stream Discharge", # graph title
       x = "Daily Rainfall (in)", # x-axis title
       y = "Average Discharge (cubic ft/s)") # y-axis title

```

$\bigstar$ **Try This:** Change the title of your graph by altering the code `title = "Linear Model of Discharge and Rainfall"`.

<br>

### If you want to get really spicy, you can even make an **interactive graph:**

This graph displays the daily rainfall and average stream discharge through time. Notice how every peak in rainfall is accompanied by a similar peak in discharge.

```{r discharge_plot, echo=FALSE, fig.height=4.5, fig.width=7}

# Create a graph:
water_plot <- ggplot(water, # uses data from the dataset "water"
       aes(x = date, # x-axis shows the date
           y = value, # y-axis shows the value of either rainfall or discharge
           fill = type)) + # fill color is determined by the type of data
  geom_area() +
  scale_fill_manual(values = c('turquoise', 'darkblue')) + # chooses colors
  facet_wrap(~type, # makes two graphs side-by-side
             ncol = 1, scales = 'free_y',
             labeller = labeller(type = labs)) +
  theme_minimal() + # sets the theme
  theme(legend.position = 'none', # removes the legend
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.spacing = unit(2, "lines")) +
  labs(title = 'Rainfall and Stream Discharge in Goleta, CA (2019)') # graph title

# Make the graph interactive:
ggplotly(water_plot) %>%
  config(displayModeBar = F) 

```

$\bigstar$ **Try This:** Interact with the graph by moving your mouse over the plotted data. You can also zoom in by clicking and dragging your cursor! Zoom out by double clicking on the plot.

<r>

### Other things R can do:

* Machine-learning
* Create web applications
* Run simulations
* Produce documents
* Make maps
* Build websites
* And much more

In fact, this tutorial was created entirely using R!

## R for EEMB 148 

Don't fret; you will not be expected to do any of these spicy and complicated things in EEMB 148. In fact, you will not be expected to generate any code at all, although you have the option of altering code that we provide. We will be focusing on learning how to to interpret data and understand the results of statistical tests.

For EEMB 148, we will use R for:

- t-tests
- ANOVAs
- Tukey tests
- Producing bar graphs

We reviewed this material in section; for reference, look at the week 2 slides from the "Biostatistics" powerpoint.

### Let's run through some code, using the same example we saw in section.

These data are from the Sierra Nevada, showing algal biomass measurements (mg/L) from three types of streams: streams with invasive brown trout, streams with declining populations of invasive brown trout, and streams with no trout.

Here are the first five rows from the dataset:

```{r algae_data, echo=FALSE}

head(algae_wider, 5)

```

Remember that each number is a measurement of algal biomass (mg/L) from a different stream.

## t-test

A t-test is used to determine whether the **means of two groups** are significantly different.

```{r t-test, exercise=TRUE}

# Run a t-test to see whether the mean of ALGAL BIOMASS varies by STREAM TYPE
# between streams with trout and without trout:
t.test(algae2$algal_biomass~algae2$stream_type)

```

$\bigstar$ **Try this:** Find the p-value and t-statistic in the output! Look at the section slides if you need help with interpretation.

## ANOVA

An ANOVA is used to determine whether the means of **three or more groups** are significantly different.

```{r ANOVA, exercise=TRUE}

# Run an ANOVA to see whether the mean of ALGAL BIOMASS varies by STREAM TYPE
# between streams with trout, with declining trout, and without trout:
algae_anova <- aov(algal_biomass ~ stream_type, data = algae) # runs the test
summary(algae_anova) # displays summary

```

$\bigstar$ **Try this:** Find the p-value in the output! It is labeled as `Pr(>F)`. Look at the section slides if you need help with interpretation.

## Tukey test

A Tukey test is used after an ANOVA, to determine **which** groups are significantly different.

```{r Tukey, exercise=TRUE}

# Run a Tukey test using the output from the ANOVA:
TukeyHSD(algae_anova)

```

$\bigstar$ **Try this:** Find the three p-values in the output! They are labeled as `P adj`, and there is one p-value for each pair of groups. Look at the section slides if you need help with interpretation.

## Bar Charts

Bar charts are the best way to visualize the output from t-tests and ANOVAs.

```{r anova_plot, exercise=TRUE, fig.width=5, fig.height=4}

# Create a summary table for plotting:
algae_summary <- algae %>% 
  group_by(stream_type) %>% 
  summarise(
    mean = mean(algal_biomass), # calculate the mean
    sd = sd(algal_biomass) # calculate standard deviation
  )

# Create a graph:
ggplot(algae_summary, # data is taken from the summary table
       aes(x = stream_type, # x-axis shows stream type
           y = mean, # y-axis shows mean algal biomass
           fill = stream_type)) + # bars are colored by stream type
  geom_col() + # graph is a bar (column) graph
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), # adds error bars using standard deviation
                width = 0.2, position = position_dodge(.9)) + # sets size of error bars
  scale_fill_manual(name = "Stream Type", # changes the name of the legend
                    values = c("blue", "orange", "darkgreen")) + # sets colors
  scale_y_continuous(expand = c(0,0)) + # removes some funky extra space
  theme_classic() + # sets the visual theme
  labs(title = "Mean Algal Biomass in Sierra Nevada Streams", # graph title
       x = "Stream Type", # x-axis title
       y ="Algal Biomass (mg/L)") # y-axis title

```

$\bigstar$ **Try this:** Change the code from `theme_classic() +` to `theme_dark()`. R has many built-in themes to choose from.

## Sources

**Congratulations, you've reached the end of this tutorial!** Wash your hands and get yourself a snack. (:

Discharge data from [USGS WaterWatch](https://waterdata.usgs.gov/nwis/dv?referred_module=sw&site_no=11120000), Site 11120000 - [Atascadero Creek](https://waterdata.usgs.gov/nwis/dv?cb_00060=on&cb_80155=on&format=rdb&site_no=11120000&referred_module=sw&period=&begin_date=2019-01-01&end_date=2019-12-31), Goleta, CA.

Rainfall data from [County of Santa Barbara Public Works rainfall data](https://rain.cosbpw.net/site.php?site_id=84&site=8339086e-5a78-4a4c-bec3-7344f97a0a80), sensor at [San Marcos Pass.](https://rain.cosbpw.net/sensor.php?time_zone=US%2FPacific&site_id=84&site=8339086e-5a78-4a4c-bec3-7344f97a0a80&device_id=2&device=44a7898b-47ed-4fee-9d83-067edccbfcd0&data_start=2019-01-01+00%3A00%3A00&data_end=2019-12-31+23%3A59%3A59&bin=86400&range=Custom+Range&legend=true&thresholds=true&refresh=off&show_raw=true&show_quality=true)

Algal biomass data courtesy of Roland Knapp.

Tutorial by Kate Culhane (kathrynculhane@ucsb.edu) and Alex Sterritt.