---
title: "EEMB 148 - HW 3" 
subtitle: "Macroinvertebrate and Fish Assessment"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: sandstone
runtime: shiny_prerendered
description: "Macroinvertebrate and fish analyses for HW 3."
---

```{r setup, include=FALSE}
# Necessary packages:

library(learnr)
library(tidyverse) 
library(agricolae)

# Necessary data:

inverts <- read_csv('macroinvertebrates.csv') %>% 
  mutate(IBI = round((coleoptera_score+ept_score+predator_score+
                        collector_score+intolerant_score+
                        noninsect_score+tolerant_score)/0.7, 1)) %>%
  mutate(site = fct_reorder(site, location, .desc = TRUE),
         location = factor(location, levels = c('upstream', 'downstream')))

fish_abundance <- read_csv('fish_abundance.csv') %>% 
  mutate(site = fct_reorder(site, location, .desc = TRUE), 
         location = factor(location, levels = c('upstream', 'downstream')))

fish_condition <- read_csv('fish_condition.csv') %>% 
  mutate(K = weight / (length^3) * 1000) %>% 
  mutate(site = factor(site, levels = c('lizard','snake','turtle',
                                        'frog','newt','salamander')), 
         location = factor(location, levels = c('upstream', 'downstream')))

# Tutorial options:

tutorial_options(exercise.timelimit = 10)

```

<style type="text/css">
.table {

    width: 500px;

}
</style>

## Study System

### San Gabriel River

The San Gabriel River flows through southern California, and is dammed in several places. The dams provide hydroelectric power, drinking water, and regulated stream flow for the Los Angeles area.

This study surveyed a stretch of the San Gabriel River near the **San Gabriel dam**, a rock-fill dam with several associated habitat alterations:

- Water is released **from the top of the dam** to maintain flow and power a hydroelectric plant.
- Native **riparian vegetation is removed** along 500 m of the river bank below dam.
- There is **substantial erosion of fine substrates** (sand and mud) into the stream below the dam.

<br>

### Study Design

A research team surveyed sites above and below the San Gabriel dam and its associated reservoir in order to quantify changes in the macroinvertebrate and fish communities. They chose **three upstream sites** within a 300-m stretch, and **three downstream sites** within a 300-m stretch. Each site was named after a type of herp that can be found in riparian habitats. "Herp" is ecological slang for reptiles and amphibians (from the word "herpetology"...no relation to STIs).

- The three sites *upstream* from the dam were named after reptiles: lizard, snake, and turtle. 
- The three sites *downstream* from the dam were named after amphibians: frog, newt, and salamander.

At each of the six sites, the research team sampled invertebrate and fish communities. **Five replicates (measurements)** were taken for each metric at each site. Therefore, each metric has $6*5=30$ total measurements.

**Experimental Design:**

- *Predictor Variable:* location (upstream or downstream of the dam) 
- *Response Variables:* invertebrate community; SoCal B-IBI; fish abundance; fish body condition (K)

<br>

### **Task:** Calculations and Statistics

Your goals are to calculate key parameters and draw statistical conclusions.

Note: All error bars represent *standard error* instead of *standard deviation*. Standard error is related to standard deviation, but takes into account sample size:

$SE=\displaystyle \frac{\sigma}{\sqrt{n}}$

Where $\sigma$ is the standard deviation and $n$ is the sample size. Like standard deviation, standard error represents the variance, or spread of the data.

***Graded Assignment:*** *Homework #3 is formatted as a "quiz" on Gauchospace.*

## INVERTEBRATE COMMUNITY 

The researchers collected invertebrate samples using the Surber sampler method, which is ideal for sampling benthic macroinvertebrates. Using these samples, the researchers measured **seven metrics** of the invertebrate community, as follows:

| Variable Name | Metric | Units | 
| :- | :- | :- | 
|`coleoptera` | Coleoptera taxa | Number of taxa |
| `ept` | EPT taxa | Number of taxa |
| `predator` | Predator taxa | Number of taxa |
| `collector` | % of collector individuals | % of all individuals | 
| `intolerant` | % of intolerant individuals | % of all individuals |
| `noninsect` | % of non-insect taxa | % of taxa | 
|`tolerant` | % of tolerant taxa | % of taxa |

Here are the raw data presented in a table. Use the black triangle in the header to view more columns. Use the green "Next" button to see more rows.

```{r invert-data, echo=FALSE}
inverts %>%
  select(-c(site_number, sample, coleoptera_score:IBI))
```

## -- Calculation: *SoCal B-IBI*

*The **So**uthern **Cal**ifornia **B**enthic Invertebrate **I**ndex of **B**iotic **I**ntegrity* (**SoCal B-IBI**) developed by Ode et al. (2005) is a comprehensive index of stream health that incorporates these seven distinct metrics. To calculate the index, you first need to score each metric on a scale from 0-10.

<br>

$\bigstar$ **Task:** The table below displays the raw data for two inverberate samples. For each sample, score the seven metrics using Table 3 from Ode et al. (2005). You will then use these scores to calculate the SoCal B-IBI. Check the section slides if you need help. *This is question #1 on the homework.*

```{r ibi-data, echo=FALSE}
inverts %>%
  filter(sample == 1, site == 'frog' | site == 'lizard') %>% 
  select(-c(site_number, sample, coleoptera_score:IBI))
```

## -- Statistics: *SoCal B-IBI*

### t-test

$\bigstar$ **Task:** Interpret the t-test output to determine whether each of the seven invertebrate metrics, as well as the SoCal B-IBI, varied by location.

The variable names for the seven metrics are: `coleoptera`, `ept`, `predator`, `collector`, `intolerant`, `noninsect`, and `tolerant`. The variable name for the SoCal B-IBI is `IBI`.

```{r ibi-ttest, exercise = TRUE}

# Run a t-test to determine whether the mean varied by location
t.test(VARIABLE ~ location,
       data = inverts)

```

<div id="ibi-ttest-hint">
**Hint:** Replace the word `VARIABLE` with one of the variable names. Press "Run Code". Find the t-statistic labeled `t` and the p-value labeled `p-value`.
</div>

<br>

### Bar graph

$\bigstar$ **Task:** Visualize the data using a bar graph. Error bars show SE (see tutorial introduction for more information).

The variable names for the seven metrics are: `coleoptera`, `ept`, `predator`, `collector`, `intolerant`, `noninsect`, and `tolerant`. The variable name for the SoCal B-IBI is `IBI`.

```{r ibi-plot, exercise=TRUE, fig.width=3, fig.height=3}

# Create a summary table for plotting:
invert_summary <- inverts %>%
  group_by(location) %>% 
  summarise(mean = mean(VARIABLE), 
            se = sd(VARIABLE)/sqrt(length(VARIABLE)))

# Create a graph using the summary table: 
ggplot(invert_summary, aes(x = location, y = mean, fill = location)) + 
  geom_col() + 
  scale_fill_manual(values = c("cadetblue", "burlywood3")) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.2, position = position_dodge(.9)) + 
  scale_y_continuous(expand = c(0,0)) + 
  theme_classic() + 
  theme(legend.position = "none") + 
  labs(title = "Mean Metric Value",
       x = "Site",
       y = "Invertebrate Metric")

```

<div id="ibi-plot-hint">
**Hint:** You need to replace `VARIABLE` in the code **three times**.
</div>

## FISH DENSITY

The **arroyo chub** (*Gila orcuttii*) is found in only a few southern California river systems, and is listed as an IUCN vulnerable species. The arroyo chub is native to the San Gabriel River system.

To sample populations of the arroyo chub, the researchers used 3-pass depletion electroshocking. After capturing all of the fish within a known area of the stream, the fish were counted as a measure of *raw abundance.* Finally, the raw abundance was divided by the *lenth of the stream sampled* in order to calculate the *density of fish.* 


| Variable Name | Metric | Units | 
| :- | :- | :- | 
| `density` | Fish density | fish / km |

Here are the raw data presented in a table.

```{r abundance-data, echo=FALSE}
fish_abundance %>% select(-c(site_number)) 
```

## -- Statistics: Fish density

### t-test

$\bigstar$ **Task:** Interpret the t-test output to determine whether the mean fish density varied by location.

```{r fish-ttest}

# Run a t-test to determine whether the mean fish density varied by location
t.test(density ~ location,
       data = fish_abundance)

```

<br>

### Bar graph

$\bigstar$ **Task:** Visualize the data using a bar graph. Error bars show SE (see tutorial introduction for more information).

```{r fish-plot, echo = FALSE, fig.width=3, fig.height=3}

# Create a summary table for plotting:
abundance_summary <- fish_abundance %>%
  group_by(location) %>% 
  summarise(mean = mean(density), 
            se = sd(density)/sqrt(length(density)))

# Create a graph using the summary table: 
ggplot(abundance_summary, aes(x = location, y = mean, fill = location)) + 
  geom_col() + 
  scale_fill_manual(values = c("cadetblue", "burlywood3")) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.2, position = position_dodge(.9)) + 
  scale_y_continuous(expand = c(0,0)) + 
  theme_classic() + 
  theme(legend.position = "none") + 
  labs(title = "Mean Density of Arroyo Chub",
       x = "Site",
       y = "Fish Density (individuals / km)")

```

## FISH CONDITION

For each arroyo chub captured, the researchers measured weight and length in order to assess fish health.


| Variable Name | Metric | Units | 
| :- | :- | :- | 
| `weight` | Weight of the fish | kg | 
| `length` | Length of the fish | cm |

Because fish abundance varied between sites, the sample sizes were not equal. This table shows sample size `n` for each site:

```{r k-summary, echo=FALSE}
fish_condition %>% 
  group_by(site) %>% 
  summarise(n = length(K))
```

Here are the raw data presented in a table.

```{r k-data, echo=FALSE} 
fish_condition %>% select(-c(site_number, sample, K))
```

## -- Calculation: *K*

We can calculate the body condition ($K$) of each fish using the equation $K=W/L^3$ where $W$ is the weight in grams and $L$ is the length in centimeters.

<br>

$\bigstar$ **Task:** Calculate $K$ using R code! Look for the phrase `condition = FORMULA` in the following code, and replace the word "FORMULA" with the correct mathematical expression.  When you click "Run Code", you should see a table with two rows of data, and a new column for the variable `K`. *This is question #6 on the homework.*

- ***Hint 1:*** *You will need to use the variables* `weight` *and* `length`*, as well as the operators* `/`*,* `^`*, and* `*`*.*
- ***Hint 2:*** *Remember that the units for* $K$ *are* $g/cm^3$*.*
- ***Hint 3:*** *If you get completely stuck, click "Solution".*

```{r calculate-K-setup}
fish_condition <- fish_condition %>% 
  filter(sample == 3,
         site == 'frog' | site == 'lizard') %>% 
  select(-c(site_number, sample, K))
```

```{r calculate-K, exercise=TRUE}

# Calculate body condition (K) using weight and length
mutate(fish_condition,
       K = FORMULA)

```

```{r calculate-K-solution}

# Calculate body condition (K) using weight and length
mutate(fish_condition,
       K = weight / (length^3) * 1000)

```

## -- Statistics: *K*

### ANOVA

$\bigstar$ **Task:** Interpret the ANOVA output to determine whether the mean fish body condition varied by site.

```{r K-ANOVA}

# Run an ANOVA to determine whether the mean fish body condition varied by site
K_anova <- aov(K ~ site, 
                 data = fish_condition)
summary(K_anova)

```

<br>

### Tukey test

$\bigstar$ **Task:** Interpret the Tukey test output to determine which sites were different.

```{r K-Tukey}

# Run a Tukey test using the output from the ANOVA
K_anova <- aov(K ~ site,
               data = fish_condition)
K_tukey <- HSD.test(K_anova, "site")
print(K_tukey$groups)

```

<br>

### Bar graph

$\bigstar$ **Task:** Visualize the data using a bar graph. Error bars show SE (see tutorial introduction for more information).

```{r K-plot, echo = FALSE, fig.width=5, fig.height=3}

# Create a summary table for plotting:
K_summary <- fish_condition %>%
  group_by(location, site) %>% 
  summarise(mean = mean(K), 
            se = sd(K)/sqrt(length(K)))

# Create a graph using the summary table: 
ggplot(K_summary, aes(x = site, y = mean, fill = location)) + 
  geom_col() + 
  scale_fill_manual(values = c("cadetblue", "burlywood3")) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.2, position = position_dodge(.9)) + 
  scale_y_continuous(expand = c(0,0)) + 
  theme_classic() + 
  theme(legend.position = "none") + 
  labs(title = "Mean Body Condtion of Arroyo Chub",
       x = "Site",
       y = "K (g/cm^3)")

```

## Sources

**Congratulations, you've reached the end of this tutorial!** Wash your hands and get yourself a snack. (:

Fish data modified from dataset courtesy of Thomas Even.

Tutorial by Kate Culhane (kathrynculhane@ucsb.edu).